The Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-08 ms.tex
blah

%% Maybe active galaxies constitute 10% need reference here?
Supermassive black holes (SMBHs) lurk in the centers of most, if not
all nearby galaxies (see reviews by,
e.g. \citealt{KormendyRichstone:1995a};
\citealt{FerrareseFord:2005a}). However, only a few percent of these
manifest themselves as luminous Active Galactic Nuclei (AGN).  Nearly
quiescent SMBHs, such as those hosting low luminosity AGN, constitute
a silent majority (e.g.~\citealt{Ho:2009a}).

To determine why most SMBHs appear to be inactive requires an understanding of their gaseous environments.  Hot gas near the sphere of influence of the SMBH (hereafter the `circumnuclear medium', or CNM) controls the mass accretion rate, $\dot{M}$.  The latter in turn sets the luminosity of the accretion flow and the impact of its energy and momentum input (`feedback') on larger physical scales.  The presence of dense gas in the nucleus may also lead to runaway cooling, resulting in bursty star-formation and AGN activity (reference Ciotti and others).

Knowledge of how $\dot{M}$ varies with the SMBH mass, $\Mbh$, and other properties of the
galactic nucleus informs key questions, such as the co-evolution of SMBHs and their host galaxies with cosmic time.  SMBH growth in the low redshift universe appears to be dominated by lower mass black holes (e.g.~\citealt{Heckman+04}).  This is often
attributed to general trend  of cosmological `down-sizing' (e.g, \citealt{Gallo+08}).  However, the physical processes involved may be distinct from those operating at higher SMBH masses and accretion rates in quasars.

The accretion rates of low mass SMBHs, and their resulting X-ray luminosities $L_X$, also have implications for what observational constraints can be placed on their occupation fraction in low mass galaxies and the possible existence of intermediate mass black holes.  By combining the average relationship between $\Mbh$ and
$L_X$ measured from a sample of high mass galaxies (\citealt{Gallo+08}) with upper limits on $L_X$ in lower mass
galaxies, \citet{MillerGallo+:2014a}  infer tentative evidence for an occupation fraction less than unity in the latter.  This method, however, relies on extrapolating a power-law fit to the $L_X-\Mbh$ distribution to low $\Mbh$ below the detection threshold.

%The conclusions of
%\citet{MillerGallo+:2014a} are thus sensitive to possible changes in
%the $L_X$-$\Mbh$ relationship at small $\Mbh$.
% AG: In our model we obtain a steeper relationship, but we have difficulty matching the observed relationship above the detection threshold

The gas density of the CNM also influences the emission from stellar tidal disruption events (TDEs), such as the high energy
transient {\it Swift} J1644+57 (\citealt{Levan+11}; \citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).
This event was powered by a transient relativistic jet, which produced
synchrotron radio emission as the jet material was decelerated by
shock interaction with the CNM of the previously quiescent SMBH
(\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed
modeling of J1644+57 showed that the CNM density was
significantly lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in J1644+57.  Variations in the properties of the CNM could thus in
principle help explain why most TDEs appear to be radio quiet
(e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

Gas comprising the CNM can in principle originate from a variety of sources: (1) wind mass loss from (predominantly evolved) stars; (2) star-star collisions; and (3) unbound debris from TDEs.  Stellar wind mass loss is probably the dominant source insofar as collisions are relevant only in very dense stellar environments for young stellar populations (\citealt{Rubin&Loeb11}), while \citet{MacLeod+13} find that TDEs contribute subdominantly to the time-averaged accretion rate of quiescent SMBHs.

@@

Another approach to determine the SMBH accretion rate, which we adopt in this paper, is to directly calculate the density, velocity and temperature profiles of the CNM using a physically motivated
hydrodynamic model.  Mass is injected to the nuclear environment via
stellar winds, while energy input results from several sources including stellar winds, supernovae, and AGN feedback (e.g., \citealt{Quataert:2004a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}).  Unlike previous works focusing on modeling individual galaxies, here we systematically analyze the properties of the CNM across a sample of galaxies with a range of SMBH masses, stellar density profiles, and star formation histories (\citealt{WangMerritt:2004a};
hereafter WM04).

@@

For fixed wind heating $\dot{M}/\dot{M}_{\rm edd}$ tends to increase
with SMBH mass $\propto M_{\bullet}^{0.5(0.8)}$ for core(cusp)
galaxies, respectively.  This appears to be in contradiction with the
general down sizing trend observed for SMBH accretion in the local
universe (e.g.~\citealt{Heckman+04}; \citealt{Gallo+08}), such as the
fact that for quiescent galaxies the nuclear X-ray luminosity obeys
$L_X \sim \Mbh^\alpha$, where $\alpha\simeq 0.7-0.8$
\citep{MillerGallo+:2014a}.  Our model instead predicts that $\Mdot
\sim \Mbh^{1.5-2}$, such that if $L_x\sim M$, then we would have $\L_x
\sim \Mbh^{1.5-2}$, steeper than the observed relationship.

@@

% This is shown in the top panel Figure~\ref{fig:bh_xray}. The
% relationship from \citealt{MillerGallo+:2014a} and their detection
% upper limits are the black and blue solid lines
% respectively. The points correspond to the $\dot{M} c^2$ for our
% solutions (scaled by $5\times 10^{-7}$).  We must choose a $\vwO$ for
% each galaxy.  We select $\vwO$=200 km/s, roughly the expected level of
% heating expected from main sequence the stellar winds and MSPs. If
% $L_x \sim \Mdot^2$ the discrepancy becomes worse, as shown in the
% bottom panel of Fig.~\ref{fig:bh_xray}, where we plot $2 \times
% 10^{-4} (\eddr) \Mdot c^2$ from our solutions with the relation from
% \citealt{MillerGallo+:2014a}.

@@

% \subsubsection{Comparison with observed X-ray Luminosity-Mass
%   Relation}
% %Note currently I reverse-engineer the stellar mass from the black
% %hole mass inferred from M-sigma; however, it would be better to use
% %the tabulated black hole mass from the Mbh-Mbulge relation.
% Fig. 1 of \citealt{MillerGallo+:2014a} plots the observed
% relationship of unresolved x-ray luminosity versus galactic stellar
% mass for a sample of quiescent galaxies. We may construct a similar
% relationship with our own sample using the inferred accretion rates
% from our solutions.  To infer an x-ray luminosity for each solution,
% we scale our calculated $\Mdot c^2$ by a constant. We pick $5\times
% 10^{-7}$, which gives x-ray luminosities comparable to
% those in \citealt{MillerGallo+:2014a}. We also try an alternate
% prescription in which $L_x$ is quadratic in $\Mdot$. In particular, we
% take $L_x=2 \times 10^{-4} (\eddr) \Mdot c^2$.

@@

% %AG-power law may be closer to 1.4 & 1.8 than to 1.5 & 2.
% Our results are plotted in Fig.~\ref{fig:bh_xray}. In our model
% there is a steeper dependence $\Mbh$ on $M_{\rm gal}$ than in
% \citealt{MillerGallo+:2014a}. We find $L_x \propto M_{\rm gal}^{1.5}$ when
% $L_x \propto M_{\rm gal}^{2}$, whereas \citealt{MillerGallo+:2014a} finds
% that $L_x \propto M_{\rm gal}^{0.7-0.8}$.

% \begin{figure}
% \includegraphics[width=\columnwidth]{bh_xray.eps}
% \caption{\label{fig:bh_xray} X-ray luminosity plotted as a function of
%   stellar mass for our solutions (points), the trend of nuclear x-ray
%   luminosity with stellar mass (black line) and the detection upper
%   limit (blue line) from Fig. 1 of \citealt{MillerGallo+:2014a}. In
%   the top panel we take the X-ray luminosity to be $\epsilon_1 \Mdot
%   c^2$, with $\epsilon_1=5\times 10^{-7}$ chosen to make our
%   luminosities comparable to those observed. In the bottom panel, we
%   take the accretion power to be quadratic in $\Mdot$. $L_x=\epsilon_2
%   \eddr \Mdot c^2$ with $\epsilon2$=$2\times10^{-4}$. For each galaxy
%   we choose the $\vwO=200$ km/s solution.  {\bf BDM: if possible, add actual x-ray data from Gallo+08!}}
% \end{figure}

>>>>

Supermassive black holes (SMBHs) lurk in the centers of most, if not
all nearby galaxies (see reviews by,
e.g. \citealt{KormendyRichstone:1995a};
\citealt{FerrareseFord:2005a}). However, only a few percent of these
manifest themselves as luminous Active Galactic Nuclei (AGN).  Nearly
quiescent SMBHs, such as those hosting low luminosity AGN, constitute
a silent majority (e.g.~\citealt{Ho:2009a}).

To determine why most SMBHs appear to be inactive requires an understanding of their gaseous environments.  Hot gas near the sphere of influence of the SMBH (hereafter the `circumnuclear medium', or CNM) controls the mass accretion rate, $\dot{M}$.  The latter in turn sets the luminosity of the accretion flow and the impact of its energy and momentum input (`feedback') on larger physical scales.  The presence of dense gas in the nucleus may also lead to runaway cooling, resulting in bursty star-formation and AGN activity (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}).

Knowledge of how $\dot{M}$ depends on the SMBH mass, $\Mbh$, and other properties of the
galactic nucleus informs key questions related to the co-evolution of SMBHs and their host galaxies with cosmic time (e.g.~\citealt{Kormendy&Ho13}; \citealt{Heckman&Best14}).  SMBH growth in the low redshift universe is dominated by low mass black holes, $M_{\bullet} \lesssim 10^{8}M_{\odot}$ (e.g.~\citealt{Heckman+04}), a fact which is often attributed to the general trend of `cosmic down-sizing' resulting from hierarchical structure growth (e.g, \citealt{Gallo+08}).  However, the physical processes involved in the growth of loss mass blacks may be distinct from those operating at higher SMBH masses and accretion rates in quasars.  One important question is whether SMBHs grow primarily by accreting gas which is fed directly from large scales, or whether some of the growth growth results from internal or secular processes such as stellar mass loss.

The accretion rates of low mass SMBHs have implications for observational constraints on the occupation fraction of SMBHs in low mass galaxies, and the potential existence of intermediate mass black holes.  \citet{Miller+15} combine the average correlation between $\Mbh$ and the nuclear X-ray luminosities, $L_{X}$ of a sample of elliptical galaxies to infer tentative evidence for a SMBH occupation fraction less than unity in galaxies with stellar masses $M_{\star} \lesssim 10^{10}M_{\odot}$ ($M_{\bullet} \lesssim 10^{7}M_{\odot}$).  This method relies on extrapolating a power-law fit of the $L_X-\Mbh$ distribution to low values of $\Mbh$ (and hence $L_{\rm X}$) below the detection threshold, despite the potential for very different physics to control the accretion rates onto the lowest mass SMBHs.

%The conclusions of
%\citet{Miller+15} are thus sensitive to possible changes in
%the $L_X$-$\Mbh$ relationship at small $\Mbh$.
% AG: In our model we obtain a steeper relationship, but we have difficulty matching the observed relationship above the detection threshold

The CNM density also influences the emission from stellar tidal disruption events (TDEs), such as the high energy
transient {\it Swift} J1644+57 (\citealt{Levan+11}; \citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).
This event was powered by a transient relativistic jet, which produced
synchrotron radio emission as the jet material was decelerated by
shock interaction with the CNM of the previously quiescent SMBH
(\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed
modeling of J1644+57 showed that the CNM density was
significantly lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in J1644+57.  Variations in the properties of the CNM could thus in
principle help explain why most TDEs appear to be radio quiet
(e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

Gas comprising the CNM can in principle originate from several sources: (1) wind mass loss from predominantly evolved stars; (2) stellar binary collisions; and (3) unbound debris from a recent TDE.  Stellar wind mass loss is probably the dominant source insofar as collisions are relevant only in extremely dense stellar environments for very young stellar populations (\citealt{Rubin&Loeb11}), while \citet{MacLeod+13} find that TDEs contribute subdominantly to the time-averaged accretion rate of quiescent SMBHs.

@@

Another approach to determine the SMBH accretion rate, which we adopt in this paper, is to directly calculate the density, velocity and temperature profiles of the CNM using a physically motivated
hydrodynamic model.  Mass is injected to the nuclear environment via
stellar winds, while energy input results from several sources including stellar winds, supernovae, and AGN feedback (\citealt{Quataert:2004a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}).  Unlike previous works focusing on modeling individual galaxies, here we systematically analyze the properties of the CNM across a sample of galaxies with a range of SMBH masses, stellar density profiles, and star formation histories (\citealt{WangMerritt:2004a}).

@@

For fixed wind heating $\dot{M}/\dot{M}_{\rm edd}$ tends to increase
with SMBH mass $\propto M_{\bullet}^{0.5(0.8)}$ for core(cusp)
galaxies, respectively.  This appears to be in contradiction with the
general down sizing trend observed for SMBH accretion in the local
universe (e.g.~\citealt{Heckman+04}; \citealt{Gallo+08}), such as the
fact that for quiescent galaxies the nuclear X-ray luminosity obeys
$L_X \sim \Mbh^\alpha$, where $\alpha\simeq 0.7-0.8$
\citep{Miller+15}.  Our model instead predicts that $\Mdot
\sim \Mbh^{1.5-2}$, such that if $L_x\sim M$, then we would have $\L_x
\sim \Mbh^{1.5-2}$, steeper than the observed relationship.

@@

% This is shown in the top panel Figure~\ref{fig:bh_xray}. The
% relationship from \citealt{Miller+15} and their detection
% upper limits are the black and blue solid lines
% respectively. The points correspond to the $\dot{M} c^2$ for our
% solutions (scaled by $5\times 10^{-7}$).  We must choose a $\vwO$ for
% each galaxy.  We select $\vwO$=200 km/s, roughly the expected level of
% heating expected from main sequence the stellar winds and MSPs. If
% $L_x \sim \Mdot^2$ the discrepancy becomes worse, as shown in the
% bottom panel of Fig.~\ref{fig:bh_xray}, where we plot $2 \times
% 10^{-4} (\eddr) \Mdot c^2$ from our solutions with the relation from
% \citealt{Miller+15}.

@@

% \subsubsection{Comparison with observed X-ray Luminosity-Mass
%   Relation}
% %Note currently I reverse-engineer the stellar mass from the black
% %hole mass inferred from M-sigma; however, it would be better to use
% %the tabulated black hole mass from the Mbh-Mbulge relation.
% Fig. 1 of \citealt{Miller+15} plots the observed
% relationship of unresolved x-ray luminosity versus galactic stellar
% mass for a sample of quiescent galaxies. We may construct a similar
% relationship with our own sample using the inferred accretion rates
% from our solutions.  To infer an x-ray luminosity for each solution,
% we scale our calculated $\Mdot c^2$ by a constant. We pick $5\times
% 10^{-7}$, which gives x-ray luminosities comparable to
% those in \citealt{Miller+15}. We also try an alternate
% prescription in which $L_x$ is quadratic in $\Mdot$. In particular, we
% take $L_x=2 \times 10^{-4} (\eddr) \Mdot c^2$.

@@

% %AG-power law may be closer to 1.4 & 1.8 than to 1.5 & 2.
% Our results are plotted in Fig.~\ref{fig:bh_xray}. In our model
% there is a steeper dependence $\Mbh$ on $M_{\rm gal}$ than in
% \citealt{Miller+15}. We find $L_x \propto M_{\rm gal}^{1.5}$ when
% $L_x \propto M_{\rm gal}^{2}$, whereas \citealt{Miller+15} finds
% that $L_x \propto M_{\rm gal}^{0.7-0.8}$.

% \begin{figure}
% \includegraphics[width=\columnwidth]{bh_xray.eps}
% \caption{\label{fig:bh_xray} X-ray luminosity plotted as a function of
%   stellar mass for our solutions (points), the trend of nuclear x-ray
%   luminosity with stellar mass (black line) and the detection upper
%   limit (blue line) from Fig. 1 of \citealt{Miller+15}. In
%   the top panel we take the X-ray luminosity to be $\epsilon_1 \Mdot
%   c^2$, with $\epsilon_1=5\times 10^{-7}$ chosen to make our
%   luminosities comparable to those observed. In the bottom panel, we
%   take the accretion power to be quadratic in $\Mdot$. $L_x=\epsilon_2
%   \eddr \Mdot c^2$ with $\epsilon2$=$2\times10^{-4}$. For each galaxy
%   we choose the $\vwO=200$ km/s solution.  {\bf BDM: if possible, add actual x-ray data from Gallo+08!}}
% \end{figure}
